Extrinsic Type Inference (DRAFT 2025-06-08)

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
organize and add notes

@@

@@

Over the past decades many type inference have been designed and analyzed 
with varying degrees of expressivity and varying degrees of intrinsic requirements
or extrinsic relaxations, with respect to both functions (e.g. functions or paths dependent on variables)
and constructors (e.g. data constructors or records).  
The first technique to infer parametric types for parametric
data without any assistance from annotations was Hindley-Milner type inference \cite{}. Despite its age, 
it remains one of the foundational 
theoretical underpinnings of practical programming languages used today, such as Standard ML \cite{}. 
For instance, in Standard ML, the system infers the identifier \code{foo} in \code{fun foo f = (f 123)}
to have type \code{forall 'a . (int -> 'a) -> 'a}
Despite its practical utility, a significant drawback of Hindley-Milner type inference is that
certain kinds of flexible programs are simply not allowed.
For instance, in Standard ML, the system rejects the program \code{fun fooPair f = (f 123, f true)}

In order to type the program \code{fooPair}, a type system would either need to determine that
\code{f} is parametric, or that the types of \code{123} and \code{true} are compatible.
Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, which
was introduced into ML-style languages using boxed polymorphism \cite{} and developed further with ML$^F$ \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
or an infinite number of
other parametric types could be used, but none is weaker than all the rest.
Therefore, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which we believe constitutes intrinsic typing of
of functions. Since there is no parameter type that is weaker than all the rest,
relying on intrinsic second order parametric type significantly restricts the reusability
of the program.

>>>>

\TODO{summarize contributions}
\TODO{full extrinsic, no annotations required to very programs on par with untyped pgograms in terms of code reusability}
\TODO{Normalizing all approximate control flow information into subtyping, allowing types to be infered in both directions}
\TODO{inductive reasoning on scalars without any intrinsic specifications}

@@

\TODO{add example of a counter stream}

@@

Our system approximates values with many forms of types
and reasons over their control with with subtyping. Our work
builds on the techniques from MLStruct \cite{} and SuperF \cite{}, which
like our system must reason over the combination of numerous forms of types.
Leading up to these systems, there have been numerous techniques developed
to reason over subsets of these type forms, with varying degrees 
of intrinsic typing.

% Over the past decades many type inference have been designed and analyzed 
% with varying degrees of expressivity and varying degrees of intrinsic requirements
% or extrinsic relaxations, with respect to both functions (e.g. functions or paths dependent on variables)
% and constructors (e.g. data constructors or records).

The first technique to infer parametric types without any assistance 
from annotations was Hindley-Milner type inference (HM)\cite{}. 
Despite its age, it remains one of the foundational 
theoretical underpinnings of practical programming languages used today, such as Standard ML \cite{}. 
For instance, in Standard ML, the system infers the identifier \code{foo} in \code{fun foo f = (f 123)}
to have type \code{forall 'a . (int -> 'a) -> 'a}
Despite its practical utility, a significant drawback of HM is that
certain kinds of general-purpose expressions are simply not allowed.
For instance, in Standard ML, the system rejects the program \code{fun fooPair f = (f 123, f true)}.

In order to type the program \code{fooPair}, a type system would either need to determine that
\code{f} is parametric, or that the types of \code{123} and \code{true} are compatible.
Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, 
which was introduced by System F \cite{}, 
introduced into ML-style languages with boxed polymorphism \cite{} 
and developed further with ML$^F$ \cite{} and various other systems \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
or an infinite number of
other parametric types could be used, but none is weaker than all the rest.
Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which constitutes intrinsic typing of
of functions. 
Since there is no parameter type that is weaker than all the rest,
merely relying on intrinsic second order parametric types 
significantly restricts the reusability of expressions.
Additionally, many of these techniques require explicit instantiation to use parametrically typed
expressions, which can be seen as an intrinsic restriction that 
delays pruning the control-flow until a particular usage.

\TODO{mention impredicative types with subtyping, bidirectional typing, local type inferenc (Pierce)}

\TODO{mention how inferring parameter types and instations is processed before and independent from their subtyping check}

\TODO{mention algebraic subtyping, Cretin, and flow analysis}

\TODO{note how solving subtyping constraints subsume the notion of bidirection typig}
